\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-28 chapters/bci.tex
Comments François, smearing figure

The key problem for heavily paralyzed patients is that the mind wants to go
where the body can't follow.
Couldn't we therefore design a system that directly interacts with the mind?
This system can then form an interface between the mind and any kind of
technology, like a robot arm or speech synthesizer, interacting with the world
on the user's behalf.
Such a device is the \ac{bci}\footnote{Sometimes also
termed brain-machine interface (BMI) when coupled to a physical actuator, like
a robot arm or an exoskeleton. The term \ac{bci} is usually preferred for
assistive technology and communication devices.}.
A \ac{bci} directly couples actions in the `real world' to the mental state of
the user.
A \ac{bci} reads your mind.

@@

\fullpagefigct{%
  \includegraphics[width=\paperwidth]{figures/bci/bci_loop.pdf}
}{%
  The \ac{bci} loop.
  The user interacts with the \ac{bci} through a \ac{bci} paradigm, in this case involving
  visual stimulation. Neural signals are recorded, and neurophysiological features
  related to the paradigm are extracted. Using machine learning, a decision can
  then be made based on these features, which can be presented back to the user.
  In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
}{%
  fig:bci/loop
}

Let's break this definition down and focus on its separate parts.
First of all, we need to identify a signal that is a direct representation of
what's going on in the brain.
Section~\ref{sec:bci-recording} gives an overview of the available options.
One might for instance measure the fluctuations in bloodflow to establish the
activity of a specific brain region
However, this signal reacts too slowly to reflect the real-time activity that is
of interest for our purposes, and it carries little information other than the
brain regions where it is originating.
A better candidate is the neuronal electrical activity.
The brain consists 86 billion neurons, which are highly interconnected cells that are
the smallest units of information processing.
The \emph{action potential} and its related \emph{postsynaptic potential}
is an electrical pulse occurring when a neuron receives
input from another neuron and is activated.
The firing of a neuron, or the combined firing of groups of
synchronized neurons, thus generate an electrical field in and near the brain,
which can be measured using electrodes.
Fluctuations in the brain's electrical potential can be faster and more
informative.

@@

The previous points to another component of a successful \ac{bci}: the specific
brain signatures related to our paradigm need to be identified within the recorded
activity.
\todo{preprocessing}
This is commonly referred to as \emph{feature extraction}
In other words, the signal needs to be interpreted \emph{in function of} the
task we want our \ac{bci} to perform.
The electrical activity of the brain is relatively weak compared to that of the
environment around us, containing electronic devices and other sources of
interference.
Furthermore, the brain is continuously processing information and carrying out
`background tasks', all of which generate their own brain activity.
On the other hand, the activity of interest in often
originates from a specific region or network within the brain.
This activity might not easily be discerned from all other brain and
environmental activity.
Compare this to trying to listen to a single speaker at a noisy party where
everyone is shouting over each other.
Some obvious sources of interference can be easily filtered out by
\emph{preprocessing}, but the problem of retrieving or \emph{decoding} only the
relevant signal is generally hard.
Nevertheless, we can attempt to solve it through supervised machine learning.
Useful information that is encoded in the brain signal by the paradigm,
can be retrieved from it by the decoder.\todomvh{the decoder needs to take into
account the control facilities of the application (degrees of freedom, serial
/parallel nature, reactive/active...).}
Section~\ref{sec:bci/decoding} takes a closer look at some common decoding
techniques.

Finally, the loop can be closed by coupling this to a
device or actuator.
There are two aspects to this.
On the one hand, the \ac{bci} gains its function by allowing the user to interact
with their environment.\todo{put focus on communication BCI}
On the other hand, the actions performed by the \ac{bci} can themselves alter
the user's cognitive state or brain activity, creating an adaptive system.
A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
Examples of this are deep brain stimulation to mitigate the symptoms of
Parkinson's disease, or the cochlear implant as a hearing prosthesis.
More indirectly, modulation can be achieved by sensory stimulation.
This involves specific, paradigm-related sensory input, like haptic
feedback\todomvh{or tactile feedback:
S.N. Flesher et al., “A brain-computer
interface that evokes tactile sensations improves robotic arm control,”
Science, vol. 372, no. 6544, pp. 831-836, 2021. } for movement \ac{bci}, or simply presenting selected actions
back to the user.
The user's brain will then adapt through reinforcement learning, causing
changes in behavior or strategy.
Neuroplasticity, the brain's ability to adapt and reinforce specific neural
pathways, can have a positive impact on \ac{bci} performance and gives rise to
rehabilitation applications.

@@

\fullpagefigct{%
  \includegraphics[width=\paperwidth]{figures/bci/recording_modalities.pdf}
}{%
	Different electrophysiology recording modalities and \\
	their respective position in a cross-section of the skull. \\
	There is a trade-off between invasiveness and  \\
	spatial resolution.
}{%
  fig:bci/recording
}
\todo{find better figure with permission to reuse and without unnecessary
modalities}

@@

Larger implanted electrodes are referred to as intracranial ac{eeg} (i\ac{eeg}).
Depth electrodes such as those used in stereo-ac{eeg} \ac{bci}~\cite{Wu2024} tasks
and closed-loop adaptive control for Deep Brain Stimulation to mitigate
Parkinson's disease symptoms~\cite{Arlotti2018}.
Since these still penetrate the cortex, electrodes that sit on top of the
cortex are more popular, as they are considered less invasive.
\Ac{ecog} is a powerful method
with many applications in \ac{bci}~\cite{Schalk2011} and epilepsy diagnosis.
Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019} with a
higher number of smaller electrodes per surface area allow for more precise
measurements.

@@

\Ac{eeg} is the de facto non-invasive electrophysiology standard in clinical
neurology practice as well as in \ac{bci} research.
Developed since 1924~\cite{Berger1929}, it is cheap and relatively practical and
offer great temporal resolution up to thousands of Hz.
\Ac{eeg} measures the electrical potential over time at a given electrode relative to a
reference electrode in volt (V), often on the scale of microvolt ($\mu$V).
As it is easily applicable on the outside of the scalp, many practical
consumer-grade \ac{eeg} headsets exist in the form (often wireless) electrode
caps or helmets.
Current consumer systems sometimes feature dry electrodes, but clinical and
research systems often use active electrodes with preamplifiers and
electrolyte gel to reduce impedance for improved signal
quality, which slightly decreases practicality, since they require a trained
individual to properly administer.
Nevertheless, it is the most accessible \ac{bci} technology for both patients
and researchers.

@@

Reactive \ac{bci} can be realized in multiple ways, depending on the
stimulation used.
Some examples include the following:
In the steady-state somatosensory evoked potentials paradigm~\cite{Petit2021},
the user can attend to one of multiple vibrotactile stimulations in different
limbs, which encodes the information of the attended limb in the brain signal.
In auditory paradigms, information can be modulated by the volume, tone, pitch,
spatial origin of presented sounds, to which the user can
attend~\cite{Kaongoen2017}.
Most common are the visual paradigms.
These can be more performant since they allows the user to use their visual
system, one of the most evolved sensory system in humans.

@@

The \ac{bci} paradigms mentioned above can also be combined to gain performance.
Straightforward examples are activating or deactivating \ac{ssvep} stimulation
with a motor command~\cite{Neeling2019}, or combining multiple visual paradigms by stimulating
along multiple dimensions at once.
\textcite{Han2023} recently used this strategy, combining frequancy and phase
coding in \ac{ssvep} with the \ac{mvep} and oddball paradigms to develop an
efficient \ac{bci} where one of 200 targets can be accurately selected with
only 800ms of stimulation.
\section{Decoders}
\label{sec:bci/decoding}

@@

\todomvh{"trial" could  be confused with intensification rounds e.g.}
\todo{remove trial alltogether}
Finally, an important concept in the evaluation of a \ac{bci} decoder, and of \acp{bci}
in general, is \ac{itr}.
\Ac{itr} reflects how fluently a \ac{bci} can be used for communication and can
be calculated as
\begin{equation}
	\text{ITR} = Q\left(\log_2N+P\log_2P+(1-P)\log_2\frac{1-P}{N-1}\right)
\end{equation}
\Ac{itr} is expressed in bits/s and is dependent on $N$, the number of
different options that can be selected per trial\footnote{Given that selections
are independent on each other. This formula requires some adaptations in the
case of sequential or hierarchical selection.}, $P$, the selection accuracy
of the decoder, and $Q$, the number of trials per second.
The parameters $N$, $P$ and $Q$ of this formula give us some insight in the
building blocks a successful, high-\ac{itr} \ac{bci}.
To improve performance, we can aim to
\begin{enumerate*}[label={\arabic*})]
\item increase $N$ by selecting a paradigm and interface that offer a
broad range of informative selections per trial,
\item increase $P$ by engineering more performant machine learning methods for
decoding, and
\item increase $Q$ by selecting a paradigm and allows fast stimulation or
responses.
\end{enumerate*}

@@

The visual oddball paradigm is a \emph{reactive}, \emph{stimulus-dependent},
\ac{bci} paradigm with all the benefits and drawbacks
Nevertheless, it can score relatively high on the \ac{itr} parameters established above.
The brain response of interest is the visual \ac{erp}, which can be accurately
measured and decoded from non-invasive \ac{eeg} signals with a relatively low
computational effort and short calibration time.
First established by~\cite{Farwell1988} in 1988, it well supported by literature
and has been proven to work for patients in day-to-day home
use~\cite{Wolpaw2018}.

\subsection{Stimulation paradigm}
One by one, visual elements on a computer screen are \emph{intensified} for a
short period of time by changing in color, brightness or size.
Intensifcations of different targets are usually 100-200ms apart and last for about
100ms~\cite{Sellers2006a}.
On each flash observed by the user, specific brain activity is elicited.
If one of these intensified elements is rare with respect to the others, i.e. it is the
\emph{odd-one out} or \emph{oddball}, this brain activity is altered.
A stimulus can be rare either due to its inherent properties like color or
location
Crucially, this is also the case if it is `marked' as rare by the user, e.g.
by paying explicit attention to only a given stimulus and not the others or
counting how many times it occurs.
If these visual elements represent letters, and we know the timing of the
intensifications of each letter, we can now establish communication by detecting
for which letter an oddball response was present in the brain signal at its
time of intensification.

\Ac{itr} in the oddball paradigm is optimized by intensifying
multiple targets at once in a row-collumnn strategy and using a sequence of
selections, giving rise to the classic matrix speller of~\cite{Farwell1988}.
Other optimizations increase response \ac{snr}, like using flashing face
images as intensifications~\cite{Jin2012} or adding distinct colors and shapes to the
stimuli~\cite{Treder2011}.
The number of targets in a visual oddball \ac{bci}is limited by the crowding
effect, which imposes a limit on how close targets can be to each other while
not distracting the user's attention~\cite{Sellers2006a,Li2010}.
This may be overcome by making hierarchical selections of stimuli representing
groups of selections~\cite{Treder2010}.

@@

The time-domain response elicited immediately after the intensification of a
stimulus is known as the visual \ac{erp}~\cite{Luck2014}.
An \ac{erp} is a waveform consisting of multiple components.
Some of these components are modulated by whether the stimulus was an
oddball or not.
These components appear as positive peaks and negative troughs in the \ac{erp}
waveform.
They follow a naming scheme based on polarity (Negative or Positive) and
latency (e.g., N2 or N200 for a negative component after $\pm200$ ms).
The component most prominently modulated by attention is called the P3
(positive deflection after $\pm300$ ms).
Therefore, oddball paradigms are frequently referred to as P300 paradigms.
However, due to the equally important contributions of other \ac{erp}
components in decoding, as we will show later, this could be considered a
misnomer.
Thus, we will adhere to the oddball paradigm naming.

@@

\paragraph{P1}
The P1 component occurs approximately 100 ms after stimulus presentation.
This component reflects initial visual processing and is primarily linked to
activity in the primary visual cortex.
P1 is particularly sensitive to gaze fixation, where the gaze is directed
toward a stimulus.
When a stimulus is fixated on, the amplitude of P1 increases, reflecting the
enhanced processing of that visual input.
In contrast, P1 shows less modulation when stimuli are attended to via
attention without direct gaze.

\paragraph{N1}
The N1 component peaks around 150 to 200 ms post-stimulus.
It represents an extension of sensory processing related to both gaze fixation
and attention.
For attended stimuli, the amplitude is greater, indicating prioritization for
further cognitive processing.
This highlights a more selective form of attention allocation, demonstrating
the impact of both types of attention on visual processing.

\paragraph{P2}
The P2 component occurs between 200 to 300 ms.
It reflects higher-order processing, particularly in contexts demanding
stimulus evaluation and feature detection.
P2 is sensitive to attention.
When participants actively engage with specific features or categories of
stimuli, P2 shows increased amplitude.
Conversely, P2 may be less responsive when gaze is not directed toward the
stimulus but attention is still maintained.

\paragraph{N2}
Peaking around 200 to 350 ms, the N2 component is associated with cognitive
control processes such as conflict monitoring and inhibition.
It is particularly pronounced in tasks requiring differentiation between
competing stimuli.
N2 reflects the allocation of attentional resources required to resolve
conflicts, irrespective of whether the gaze is directly on the competing
stimuli or not.
Higher amplitudes in N2 are seen when cognitive demands increase, indicating
the influence of attention strategies.

\paragraph{P3}
The P3 component is elicited in oddball paradigms and peaks around 300 ms
post-stimulus.
It reflects attentional engagement and the processing of rare or unexpected
events.
The P3 is typically subdivided into two subcomponents: P3a and P3b.
The P3a is associated with the allocation of attention to novel or unexpected
stimuli, indicating the initial detection of a change in the environment.
It is particularly responsive to stimuli that capture attention, whether
through gaze fixation or attention.
P3a typically exhibits a frontal distribution and peaks earlier than P3b.

@@

\paragraph{Late Positive Potential (LPP)}
The LPP is a later component occurring beyond 400 ms.
Sustained attention and emotional processing of stimuli are reflected in the
LPP.
Typically enhanced for emotionally salient or motivationally relevant
information, LPP serves as an indicator of cognitive appraisal beyond mere
recognition tasks.
This component can be influenced by attention when the participant evaluates
the emotional significance of stimuli, even if gaze is not directed at them.

In summary, ERP components such as P1, N1, P2, N2, P3, and LPP collectively
elucidate the neural underpinnings of visual processing and attentional
mechanisms.
These components not only inform the design of an effectiv oddball BCI system
but also enhance our understanding of the cognitive processes involved in visual
attention.
This includes both gaze fixation and attention.
Understanding these components is essential for optimizing BCI protocols and
improving user interaction, particularly in contexts involving visual oddball
tasks.

>>>>

The key problem for heavily paralyzed patients is that the mind wants to go
where the body can't follow.
Couldn't we therefore design a system that directly interacts with the mind?
This system can then form an interface between the mind and any kind of
technology, like a robot arm or speech synthesizer, interacting with the world
on the user's behalf.
Such a device is the \ac{bci}\footnote{Sometimes also
termed brain-machine interface (BMI) when coupled to a physical actuator, like
a robot arm or an exoskeleton. The term \ac{bci} is usually preferred for
assistive technology and communication devices.}.
A \ac{bci} directly couples actions in the `real world' to the mental state of
the user.
%A \ac{bci} reads your mind.

@@

\fullpagefigct{%
  \includegraphics[width=\textwidth]{figures/bci/bci_loop.pdf}
}{%
  The \ac{bci} loop.
  The user interacts with the \ac{bci} via a specific paradigm, in this case involving
  visual stimulation.
  Electrical neural signals are recorded, and neurophysiological features
  related to the paradigm are extracted. Using machine learning, a decision can
  then be made based on these features, which can be presented back to the user.
  In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
}{%
  fig:bci/loop
}

Let's break this definition down and focus on its separate parts.
First of all, we need to identify a signal that is a direct representation of
what's going on in the brain.
Section~\ref{sec:bci-recording} gives an overview of the available options.
One might for instance measure the fluctuations in bloodflow to establish the
activity of a specific brain region
However, this signal reacts too slowly to reflect the real-time activity that is
of interest for our purposes, and it carries little information other than the
brain regions where it is originating.
A better candidate is the neuronal electrical activity.
The brain contains 86 billion neurons, which are highly interconnected cells that are
the smallest units of information processing.
The \emph{action potential} and its related \emph{postsynaptic potential}
is an electrical pulse occurring when a neuron receives
input from another neuron and is activated.
The firing of a neuron, or the combined firing of groups of
synchronized neurons, thus generate an electrical field in and near the brain,
which can be measured using electrodes.
Fluctuations in the brain's electrical activity are be faster and more
informative.

@@

The previous points to another component of a successful \ac{bci}: the specific
brain signatures related to our paradigm need to be identified within the recorded
activity.
\todo{preprocessing}
This is commonly referred to as \emph{feature extraction}.
In other words, the signal needs to be interpreted \emph{in function of} the
task we want our \ac{bci} to perform.
The electrical activity of the brain is relatively weak compared to that of the
environment around us, containing electronic devices and other sources of
interference.
Furthermore, the brain is continuously processing information and carrying out
`background tasks', all of which generate their own brain activity.
On the other hand, the activity of interest often
originates from a specific region or network within the brain.
This activity might not easily be discerned from all other brain and
environmental activity.
Compare this to trying to listen to a single speaker at a noisy party where
everyone is shouting over each other.
Some obvious sources of interference can be easily filtered out by
\emph{preprocessing}, but the problem of retrieving or \emph{decoding} only the
relevant signal is generally hard.
Nevertheless, we can attempt to solve it through supervised machine learning.
Useful information that is encoded in the brain signal by the paradigm,
can be retrieved from it by the decoder.\todomvh{the decoder needs to take into
account the control facilities of the application (degrees of freedom, serial
/parallel nature, reactive/active...).}
Section~\ref{sec:bci/decoding} takes a closer look at some common decoding
techniques.

Finally, the loop can be closed by coupling this to a
device or actuator.
There are two aspects to this.
On the one hand, the \ac{bci} gains its function by allowing the user to interact
with their environment.\todo{put focus on communication BCI}
On the other hand, the actions performed by the \ac{bci} can themselves alter
the user's cognitive state or brain activity, creating an adaptive system.
A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
Examples of this are deep brain stimulation to mitigate the symptoms of
Parkinson's disease, or the cochlear implant as a hearing prosthesis.
More indirectly, modulation can be achieved by sensory stimulation.
This involves specific, paradigm-related sensory input, like haptic
feedback\todomvh{or tactile feedback:
S.N. Flesher et al., “A brain-computer
interface that evokes tactile sensations improves robotic arm control,”
Science, vol. 372, no. 6544, pp. 831-836, 2021. } for movement \ac{bci}, or simply presenting selected actions
back to the user.
The user's brain will then adapt through reinforcement learning, causing
changes in behavior or strategy.
Neuroplasticity, the brain's ability to adapt and reinforce specific neural
pathways, can have a positive impact on \ac{bci} performance and gives rise to
rehabilitation applications.
\todo{is every section referred to?}

@@

\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/bci/recording_modalities.pdf}
}{%
	Different electrophysiology recording modalities and \\
	their respective position in a cross-section of the skull. \\
	There is a trade-off between invasiveness and  \\
	spatial resolution.
}{%
  fig:bci/recording
}
\todo{find better figure with permission to reuse and without unnecessary
modalities}

@@

Larger implanted electrodes are referred to as intracranial \ac{eeg} (i\ac{eeg}).
Depth electrodes such as those used in stereo-ac{eeg} \ac{bci}~\cite{Wu2024} tasks
and closed-loop adaptive control for Deep Brain Stimulation to mitigate
Parkinson's disease symptoms~\cite{Arlotti2018}.
Since these still penetrate the cortex, electrodes that sit on top of the
cortex are more popular, as they are considered less invasive.
\Ac{ecog} is a powerful method
with many applications in \ac{bci}~\cite{Schalk2011} and epilepsy diagnosis.
Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019} with a
higher number of smaller electrodes per surface area allow for more precise
measurements.

@@

\Ac{eeg} is the de facto non-invasive electrophysiology standard in clinical
neurology practice as well as in \ac{bci} research.
Developed since 1924~\cite{Berger1929}, it is cheap and relatively practical and
offers great temporal resolution up to thousands of Hz.
\Ac{eeg} measures the electrical potential over time at a given electrode relative to a
reference electrode in volt (V), often on the scale of microvolt ($\mu$V).
As it is easily applicable on the outside of the scalp, many practical
consumer-grade \ac{eeg} headsets exist in the form of (often wireless) electrode
caps or helmets.
Current consumer systems sometimes feature dry electrodes, but clinical and
research systems often use active electrodes with preamplifiers and
electrolyte gel to reduce impedance for improved signal
quality, which slightly decreases practicality, since they require a trained
individual to properly administer.
Nevertheless, it is the most accessible \ac{bci} technology for both patients
and researchers.

@@

Reactive \ac{bci} can be realized in multiple ways, depending on the
stimulation used.
Some examples include the following:
In the steady-state somatosensory evoked potentials paradigm~\cite{Petit2021},
the user can attend to one of multiple vibrotactile stimulations in different
limbs, which encodes the information of the attended limb in the brain signal.
In auditory paradigms, information can be modulated by the volume, tone, pitch,
spatial origin of presented sounds, to which the user can
attend~\cite{Kaongoen2017}.
Most common are the visual paradigms.
These can be more performant since they allow the user to use their visual
system, one of the most evolved sensory system in humans.

@@

The \ac{bci} paradigms mentioned above can also be combined to gain performance.
Straightforward examples are activating or deactivating \ac{ssvep} stimulation
with a motor command~\cite{Neeling2019}, or combining multiple visual paradigms by stimulating
along multiple dimensions at once.
\textcite{Han2023} recently used this strategy, combining frequency and phase
coding in \ac{ssvep} with the \ac{mvep} and oddball paradigms to develop an
efficient \ac{bci} where one of 200 targets can be accurately selected with
only 800ms of stimulation.
\section{Decoders}
\label{sec:bci/decoding}

@@

\todomvh{"trial" could  be confused with intensification rounds e.g.}
\todo{remove trial alltogether}
Finally, an important concept in the evaluation of a \ac{bci} decoder, and of \acp{bci}
in general, is \ac{itr}.
\Ac{itr} reflects how fluently a \ac{bci} can be used for communication and can
be calculated as
\begin{equation}
	\text{ITR} = Q\left(\log_2N+P\log_2P+(1-P)\log_2\frac{1-P}{N-1}\right)
\end{equation}
\Ac{itr} is expressed in bits/s and is dependent on $N$, the number of
different options that can be selected per trial\footnote{Given that selections
are independent on each other. This formula requires some adaptations in the
case of sequential or hierarchical selection.}, $P$, the selection accuracy
of the decoder, and $Q$, the number of trials per second.
The parameters $N$, $P$ and $Q$ of this formula give us some insight in the
building blocks a successful, high-\ac{itr} \ac{bci}.
To improve performance, we can aim to
\begin{enumerate*}[label={\arabic*})]
\item increase $N$ by selecting a paradigm and interface that offer a
broad range of informative selections per trial,
\item increase $P$ by engineering more performant machine learning methods for
decoding, and
\item increase $Q$ by selecting a paradigm that allows fast stimulation or
responses.
\end{enumerate*}

@@

The visual oddball paradigm is a \emph{reactive}, \emph{stimulus-dependent},
\ac{bci} paradigm with all the benefits and drawbacks.
Nevertheless, it can score relatively high on the \ac{itr} parameters established above.
The brain response of interest is the visual \ac{erp}, which can be accurately
measured and decoded from non-invasive \ac{eeg} signals with a relatively low
computational effort and short calibration timeand .
First established by~\cite{Farwell1988} in 1988, it is well supported by literature
and has been proven to work for patients in day-to-day home
use~\cite{Wolpaw2018}.

\subsection{Stimulation paradigm}
One by one, visual elements on a computer screen are \emph{intensified} for a
short period of time by changing in color, brightness or size.
Intensifcations of different targets are usually 100-200ms apart and last for about
100ms~\cite{Sellers2006a}.
On each flash observed by the user, specific brain activity is elicited.
If one of these intensified elements is rare with respect to the others, i.e. it is the
\emph{odd-one out} or \emph{oddball}, this brain activity is altered.
A stimulus can be rare either due to its inherent properties like color or
location.
Crucially, this is also the case if it is `marked' as rare by the user, e.g.
by paying explicit attention to only a given stimulus and not the others or
counting how many times it occurs.
If these visual elements represent letters, and we know the timing of the
intensifications of each letter, we can now establish communication by detecting
for which letter an oddball response was present in the brain signal at its
time of intensification.

\Ac{itr} in the oddball paradigm is optimized by intensifying
multiple targets at once in a row-collumn strategy and using a sequence of
selections, giving rise to the classic matrix speller of~\cite{Farwell1988}.
Other optimizations increase response \ac{snr}, like using flashing face
images as intensifications~\cite{Jin2012} or adding distinct colors and shapes to the
stimuli~\cite{Treder2011}.
The number of targets in a visual oddball \ac{bci} is limited by the crowding
effect, which imposes a limit on how close targets can be to each other while
not distracting the user's attention~\cite{Sellers2006a,Li2010}.
This may be overcome by making hierarchical selections of stimuli representing
groups of selections~\cite{Treder2010}.

@@

The time-domain response elicited immediately after the intensification of a
stimulus is known as the visual \ac{erp}~\cite{Luck2014}.
An \ac{erp} is a waveform consisting of multiple components.
Some of these components are modulated by whether the stimulus was an
oddball or not.
These components appear as positive peaks and negative troughs in the \ac{erp}
waveform.
They follow a naming scheme based on polarity (Negative or Positive) and
latency (e.g., N2 or N200 for a negative component after $\pm200$ ms).
The component most prominently modulated by attention is called the P3
(positive deflection after $\pm300$ ms).\todo{add footnote about how ms naming
is maybe not suited, especially since this work focuses on the fact that there
is a lot of variability in the timings for visual potentials. the short naming
scheme builds on 'the first negative component', 'the first positive
component', and for the visual oddball erp this ordening more or less coincides
with the timings.}
Therefore, oddball paradigms are frequently referred to as P300 paradigms.
However, due to the equally important contributions of other \ac{erp}
components in decoding, as we will show later, this could be considered a
misnomer.
Thus, we will adhere to the oddball paradigm naming.

@@

\emph{The P1 component} occurs approximately 100 ms after stimulus presentation.
This component reflects initial visual processing and is primarily linked to
activity in the primary visual cortex.
P1 is particularly sensitive to gaze fixation, where the gaze is directed
toward a stimulus.
When a stimulus is fixated on, the amplitude of P1 increases, reflecting the
enhanced processing of that visual input.
In contrast, P1 shows less modulation when stimuli are attended to via
attention without direct gaze.

\emph{The N1 component} peaks around 150 to 200 ms post-stimulus.
It represents an extension of sensory processing related to both gaze fixation
and attention.
For attended stimuli, the amplitude is greater, indicating prioritization for
further cognitive processing.
This highlights a more selective form of attention allocation, demonstrating
the impact of both types of attention on visual processing.

\emph{The P2 component} occurs between 200 to 300 ms.
It reflects higher-order processing, particularly in contexts demanding
stimulus evaluation and feature detection.
P2 is sensitive to attention.
When participants actively engage with specific features or categories of
stimuli, P2 shows increased amplitude.
Conversely, P2 may be less responsive when gaze is not directed toward the
stimulus but attention is still maintained.

\emph{The N2 component} peaking around 200 to 350 ms is associated with cognitive
control processes such as conflict monitoring and inhibition.
It is particularly pronounced in tasks requiring differentiation between
competing stimuli.
N2 reflects the allocation of attentional resources required to resolve
conflicts, irrespective of whether the gaze is directly on the competing
stimuli or not.
Higher amplitudes in N2 are seen when cognitive demands increase, indicating
the influence of attention strategies.

\emph{The P3 component} is elicited in oddball paradigms and peaks around 300 ms
post-stimulus.
It reflects attentional engagement and the processing of rare or unexpected
events.
The P3 is typically subdivided into two subcomponents: P3a and P3b.
The P3a is associated with the allocation of attention to novel or unexpected
stimuli, indicating the initial detection of a change in the environment.
It is particularly responsive to stimuli that capture attention, whether
through gaze fixation or attention.
P3a typically exhibits a frontal distribution and peaks earlier than P3b.

@@

\emph{The late positive potential (LPP)} is a later component occurring beyond 400 ms.
Sustained attention and emotional processing of stimuli are reflected in the
LPP.
Typically enhanced for emotionally salient or motivationally relevant
information, the LPP serves as an indicator of cognitive appraisal beyond mere
recognition tasks.
This component can be influenced by attention when the participant evaluates
the emotional significance of stimuli, even if gaze is not directed at them.

In summary, ERP components such as P1, N1, P2, N2, P3, and LPP collectively
elucidate the neural underpinnings of visual processing and attentional
mechanisms.
These components not only inform the design of an effective oddball BCI system
but also enhance our understanding of the cognitive processes involved in visual
attention.
This includes both gaze fixation and attention.
Understanding these components is essential for optimizing BCI protocols and
improving user interaction, particularly in contexts involving visual oddball
tasks.
